---
title: "Section4. Meta_analysis"
output:
  html_document: default
  pdf_document: default
date: "2024-07-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This section provides a guide on how to perform meta-analysis to summarize the effects of the sex onto overall brain tissues. <br>

<br>
<br>

#### Import the expression data (Beta and SE from random effects)

```{r Importe_expression, echo=TRUE}
association_file=list.files(path = "/Users/muyangzhang/sex_research_outputs/brain_genes_imrob/brain_regions_lmrob_rmd/Brain_Regions_Imrob_Sex_outputs_txt")
association_file
```


```{r Importe_expression-2, echo=TRUE}
###Visualize the files in the "../../Data/TPM_tissue" folder

data=read.table(paste("/Users/muyangzhang/sex_research_outputs/brain_genes_imrob/brain_regions_lmrob_rmd/Brain_Regions_Imrob_Sex_outputs_txt/",association_file[1],sep=''),
                stringsAsFactors = F,header=T)
ID=unlist(strsplit(association_file[1],'_'))[2]
data=data[!is.na(data$Pr...t..),]
data=data[,1:3]

names(data)=c("Gene",paste(ID,"_Beta",sep=''),paste(ID,"_SE",sep=''))

for (i in 2:length(association_file))
{
  newdata=read.table(paste("/Users/muyangzhang/sex_research_outputs/brain_genes_imrob/brain_regions_lmrob_rmd/Brain_Regions_Imrob_Sex_outputs_txt/",association_file[i],sep=''),stringsAsFactors = F,header=T)
  
  ID=unlist(strsplit(association_file[i],'_'))[2]
  newdata=newdata[!is.na(newdata$Pr...t..),]
  newdata=newdata[,1:3]
  names(newdata)=c("Gene",paste(ID,"_Beta",sep=''),paste(ID,"_SE",sep=''))

  data=merge(data,newdata,by='Gene',all=T)
}


```

### Keep genes that detected in at least two tissues

```{r clean_data, echo=TRUE}

data_row_na=apply(data,1,function(x) sum(is.na(x)))

data=data[data_row_na<24,]

```

<br>
<br>

#### 





```{r meta_analysis-1, echo=TRUE}
library(metafor)
library(foreach)
library(doParallel)
```


```{r meta_analysis-2, echo=TRUE}
meta_test=function(data,GENE_ID)
{
  beta=as.vector(unlist(data[data$Gene==GENE_ID,grepl("_Beta",names(data))]))
  se=as.vector(unlist(data[data$Gene==GENE_ID,grepl("_SE",names(data))]))
  
  test=rma(yi=beta,sei=se,method='REML',control=list(maxiter=5000,stepadj=0.5))
  
  return(c(GENE_ID,test$k.eff,test$beta,test$se,test$zval,test$pval,test$I2,test$H2,test$QE,test$QEp))
}

# Set the number of cores you want to use
num_cores <- detectCores()

# Register parallel backend
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Initialize an empty list to store the results
results_list <- foreach(i = 1:dim(data)[1], .combine = rbind, .packages='metafor') %dopar% {
  # Calculate each result in parallel
  meta_test(data, data$Gene[i])
}

# Close the cluster
stopCluster(cl)

# Convert the list of results to a data frame
results <- as.data.frame(results_list)

names(results)=c("gene_ID","RE_sample_size",
                 "RE_beta","RE_SE","RE_Z","RE_metaP",
                 "RE_I2","RE_H2","RE_Q_val","RE_Q_pval")

results$RE_metaP_FDR=p.adjust(as.numeric(results$RE_metaP),method='fdr')

results=results[order(results$RE_metaP_FDR),]

sum(as.numeric(results$RE_metaP_FDR)<0.05)
sum(as.numeric(results$RE_metaP)<0.05)

write.table(results,"brain_13_region_sex_meta_results.txt",quote=F,sep='\t',row.names = F)

```

<br>
<br>

### QC on meta-analysis data - number of detected tissues

```{r meta_QC_number, echo=TRUE}

table(results$RE_sample_size)
table(results$RE_sample_size,as.numeric(results$RE_metaP_FDR)<0.05)

```

> There are some genes only detected in one brain region, so attention whould be required when performing the downstream analysis

<br>
<br>


### QC on meta-analysis data - number of detected tissues and their significance

```{r meta_QC_number_FDR, echo=TRUE}

table(results$RE_sample_size,results$RE_metaP_FDR<0.05)


```
> we see that majority of significant genes were detected in all 13 brain regions. Those set can be focused in some of downstream analysis


<br>
<br>

### QC on meta-analysis data - heterogeneity

```{r meta_QC_number_FDR-2, echo=TRUE}

table(results$RE_sample_size[results$RE_metaP_FDR<0.05],(results$RE_Q_pval>0.05)[results$RE_metaP_FDR<0.05])


```


### GO analysis of the results (all genes) with low heterogenity.

```{r GO_analysis, echo=TRUE}

library(gprofiler2)

gene_ID_clean=as.vector(sapply(results$gene_ID,function(x) unlist(strsplit(x,'[.]'))[1]))

gostres2 <- gost(query = c(gene_ID_clean[results$RE_metaP_FDR<0.05 & results$RE_Q_pval > 0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_results=as.matrix(gostres2$result)
GO_results=GO_results[order(as.numeric(unlist(GO_results[,3]))),]

write.table(GO_results,"brain_13_region_Sex_GO_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_results)
```

<br>
<br>

### GO analysis of the up regulated genes with low heterogenity.

```{r GO_analysis_UP, echo=TRUE}

library(gprofiler2)

gostres2_UP <- gost(query = c(gene_ID_clean[results$RE_metaP_FDR<0.05 & results$RE_beta>0 & results$RE_Q_pval>0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_UP_results=as.matrix(gostres2_UP$result)
GO_UP_results=GO_UP_results[order(as.numeric(unlist(GO_UP_results[,3]))),]

write.table(GO_UP_results,"brain_13_region_Sex_GO_UP_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_UP_results)

```


<br>
<br>

### GO analysis of the down regulated genes with low heterogenity

```{r GO_analysis_down, echo=TRUE}

library(gprofiler2)

gostres2_DOWN <- gost(query = c(gene_ID_clean[results$RE_metaP_FDR<0.05 & results$RE_beta<0 & results$RE_Q_pval>0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_DOWN_results=as.matrix(gostres2_DOWN$result)
GO_DOWN_results=GO_DOWN_results[order(as.numeric(unlist(GO_DOWN_results[,3]))),]

write.table(GO_DOWN_results,"brain_13_Sex_region_GO_DOWN_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_DOWN_results)

```
<br>
<br>

### Extracting the names of Genes.
```{r Extracting the Gene_ID}
gene_ID_list <- results$gene_ID
write.table(gene_ID_list, "gene_ID_list.txt", quote = FALSE, row.names = FALSE, col.names = FALSE)
```
```{r}
head(results)
```

### Importing the Chromsome Data
```{r Importing the chromsome data}
chromosome_data <- read.table("/Users/muyangzhang/Downloads/mart_export.txt", header = TRUE, sep = "\t")
colnames(chromosome_data) <- c("gene_ID", "Chromosome", "Gene_Start", "Gene_End")
head(chromosome_data)
```
### Merging the datasets
```{r Merging the datasets}
results <- merge(results, chromosome_data, by = "gene_ID")
head(results)
```


### Cagatorize the chromosome data based on sex and non-sex
```{r Cagatorize the chromsome data based on sex and non-sex}
results$Chromosome_type = ifelse(results$Chromosome %in% c("X", "Y"), "Sex", "Autosomal")
head(results)
```
```{r Saving the Meta_analysis_results_with Chromosome Annotated}
write.table(results,"brain_13_Sex_region_meta_Chromsome_Annotated.txt",quote=F,sep='\t',row.names = F)
```
### Seperate the data based on the charomsome type
```{r Seperate the data based on the charomsome type}
sex_results <- subset(results, Chromosome_type == "Sex")
auto_results <- subset(results, Chromosome_type == "Autosomal")
```
```{r}
head(sex_results)
head(auto_results)
```
### FDR Adjustment
```{r FDR adjustment}
sex_results$RE_metaP_FDR_Adjusted <- p.adjust(sex_results$RE_metaP, method = "fdr")
auto_results$RE_metaP_FDR_Adjusted <- p.adjust(auto_results$RE_metaP, method = "fdr")
```
```{r}
head(sex_results)
```

### Volcano plot of meta-analysis results - Sex results with low heterogenity

```{r volcano plot-1, echo=TRUE, fig.margin=TRUE}
library(ggplot2)

volcano_data=sex_results[,c("RE_beta","RE_metaP","RE_metaP_FDR_Adjusted", "RE_Q_pval")]

volcano_data$color[volcano_data$RE_Q_pval>0.05] <- "gray"
volcano_data$color[volcano_data$RE_metaP_FDR_Adjusted<0.05 & volcano_data$RE_beta>0 & volcano_data$RE_Q_pval>0.05]='red'
volcano_data$color[volcano_data$RE_metaP_FDR_Adjusted<0.05 & volcano_data$RE_beta<0 & volcano_data$RE_Q_pval>0.05]='blue'



ggplot(data=volcano_data, aes(x=as.numeric(RE_beta), y=-log10(as.numeric(RE_metaP)), col=color)) +
        geom_point() + 
        theme_minimal() +
        scale_color_manual(values=c("blue", "grey80", "red"))+
  theme_classic()

```
### Volcano plot of meta-analysis results - Non-Sex results with low heterogenity
```{r volcano plot-2, echo=TRUE, fig.margin=TRUE}
volcano_data=auto_results[,c("RE_beta","RE_metaP","RE_metaP_FDR_Adjusted", "RE_Q_pval")]

volcano_data$color[volcano_data$RE_Q_pval>0.05] <- "gray"
volcano_data$color[volcano_data$RE_metaP_FDR_Adjusted<0.05 & volcano_data$RE_beta>0 & volcano_data$RE_Q_pval>0.05]='red'
volcano_data$color[volcano_data$RE_metaP_FDR_Adjusted<0.05 & volcano_data$RE_beta<0 & volcano_data$RE_Q_pval>0.05]='blue'



ggplot(data=volcano_data, aes(x=as.numeric(RE_beta), y=-log10(as.numeric(RE_metaP)), col=color)) +
        geom_point() + 
        theme_minimal() +
        scale_color_manual(values=c("blue", "grey80", "red"))+
  theme_classic()
```


### display the key featured genes (sex) - ones with extremely low meta p-values

```{r check_p_value, echo=TRUE, fig.margin=TRUE}

sex_results[as.numeric(sex_results$RE_metaP)<(1e-50),]

```
### display the key featured genes (non-sex) - ones with extremely low meta p-values
```{r check_p_value-2, echo=TRUE, fig.margin=TRUE}
auto_results[as.numeric(auto_results$RE_metaP)<(1e-50),]
```
### GO_analysis_non_sex with low heterogenity
```{r GO_analysis_non_sex, echo=TRUE}

library(gprofiler2)

gene_ID_clean=as.vector(sapply(auto_results$gene_ID,function(x) unlist(strsplit(x,'[.]'))[1]))

gostres2 <- gost(query = c(gene_ID_clean[auto_results$RE_metaP<0.05 & auto_results$RE_Q_pval > 0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_results=as.matrix(gostres2$result)
GO_results=GO_results[order(as.numeric(unlist(GO_results[,3]))),]

write.table(GO_results,"brain_13_region_Non_Sex_GO_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_results)
```

### GO_analysis_non_sex with low heterogenity(Upregulated)
```{r GO_analysis_UP_non_sex, echo=TRUE}

library(gprofiler2)

gostres2_UP <- gost(query = c(gene_ID_clean[auto_results$RE_metaP<0.05 & auto_results$RE_beta>0 & auto_results$RE_Q_pval>0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_UP_results=as.matrix(gostres2_UP$result)
GO_UP_results=GO_UP_results[order(as.numeric(unlist(GO_UP_results[,3]))),]

write.table(GO_UP_results,"brain_13_region_Non_Sex_GO_upregulated_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_UP_results)

```

### GO_analysis_non_sex with low heterogenity(Down_regulated)
```{r GO_analysis_down_non_sex, echo=TRUE}

library(gprofiler2)

gostres2_UP <- gost(query = c(gene_ID_clean[auto_results$RE_metaP<0.05 & auto_results$RE_beta<0 & auto_results$RE_Q_pval>0.05]), 
                organism = "hsapiens", user_threshold = 0.05, correction_method = "fdr", 
                domain_scope = "annotated")

GO_UP_results=as.matrix(gostres2_UP$result)
GO_UP_results=GO_UP_results[order(as.numeric(unlist(GO_UP_results[,3]))),]

write.table(GO_UP_results,"brain_13_region_Non_Sex_GO_downregulated_results_FDR0.05.txt",quote=F,sep='\t',row.names = F)

head(GO_UP_results)

```
<br>

>> the top candidates are from sex chromosome, thus, proper analsyis should be conducted to stratify results from sex or non-sex chromosome
