---
title: "Brain. QC_and_multithreading"
output:
  html_document: default
  pdf_document: default
date: "2024-07-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



### Read the expression, phenotype, and covariate data into workspace

<br>

#### Import the expression data 
```{r Importe_expression, echo=TRUE}
brain_cortex_tmp=read.table("/Users/muyangzhang/Desktop/research_sex/gtxforr/tpm_tissue/gene_tpm_brain_cortex.gct",stringsAsFactors = F,header=T,skip=2,check.names = F,sep='\t')

brain_cortex_tmp_clean=brain_cortex_tmp[,c(2,4:dim(brain_cortex_tmp)[2])]

```

<br>
<br>

#### Reformat the expression data 
```{r Reformat_expression, echo=TRUE}

brain_cortex_tmp_new=as.data.frame(t(brain_cortex_tmp_clean))
names(brain_cortex_tmp_new)=brain_cortex_tmp_new[1,]
brain_cortex_tmp_new=brain_cortex_tmp_new[-1,]
brain_cortex_tmp_new=cbind(data.frame(Tissue.ID=row.names(brain_cortex_tmp_new),
                               Individual.ID=unlist(sapply(row.names(brain_cortex_tmp_new),function(x) sub("^([^\\-]+\\-[^\\-]+).*", "\\1", x)))),brain_cortex_tmp_new)
head(brain_cortex_tmp_new)
```

<br>
<br>

### Import the phenotype data

```{r Importe_phenotype, echo=TRUE}

phenotype=read.table("/Users/muyangzhang/Desktop/research_sex/gtxforr/metadata_files/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt", stringsAsFactors = F,header=T,check.names = F,sep='\t')
head(phenotype)

```

<br>
<br>

### Import covariate data into R

```{r Importe_cov, echo=TRUE}

brain_cortex_cov=read.table("/Users/muyangzhang/Desktop/research_sex/gtxforr/cis_qtl/GTEx_Analysis_v8_eQTL_covariates/Brain_Cortex.v8.covariates.txt",stringsAsFactors = F,header=T,check.names = F,sep='\t')

brain_cortex_cov_new=as.data.frame(t(brain_cortex_cov))
names(brain_cortex_cov_new)=brain_cortex_cov_new[1,]
brain_cortex_cov_new=brain_cortex_cov_new[-1,]
brain_cortex_cov_new=cbind(data.frame(Individual.ID=row.names(brain_cortex_cov_new)),brain_cortex_cov_new)

```


#### Merge the three datasets

```{r Merge_TPM_pheno, echo=TRUE}

brain_cortex_tmp_pheno=merge(brain_cortex_tmp_new,phenotype,by.x='Individual.ID',by.y='SUBJID')
brain_cortex_tmp_pheno_cov=merge(brain_cortex_tmp_pheno,brain_cortex_cov_new,by='Individual.ID')

```


<br>
<br>

#### converting age group to middle value
```{r change_age, echo=TRUE, fig.margin=TRUE}
library(dplyr)

values_to_replace <- unique(brain_cortex_tmp_pheno_cov$AGE)
```


```{r change_age, echo=TRUE, fig.margin=TRUE}
values_to_replace=values_to_replace[order(values_to_replace)]

replacement_values <- c(25,35,45,55,65,75)

brain_cortex_tmp_pheno_cov$AGE = sapply(brain_cortex_tmp_pheno_cov$AGE, function(x)
                                 replacement_values[which(x==values_to_replace)])
head(brain_cortex_tmp_pheno_cov)
```

<br>
<br>

#### Examine mean expression of all detected genes

```{r calculate_mean, echo=TRUE, fig.margin=TRUE}

gene_col_ID=which(grepl("ENSG",names(brain_cortex_tmp_pheno_cov)))

gene_mean=data.frame(geneID=names(brain_cortex_tmp_pheno_cov)[gene_col_ID],
                     mean_TPM=apply(brain_cortex_tmp_pheno_cov[,gene_col_ID],2,function(x) mean(as.numeric(x))))

hist(gene_mean$mean_TPM,breaks=200000,xlim=c(0,10))
```
<br>
<br>

#### check number of genes with mean TPM < 1 in blood

```{r check_TPM, echo=TRUE, fig.margin=TRUE}

table(gene_mean$mean_TPM<1)
```
> There are 39374 genes with TPM < 1, which can be filtered during the analysis.

<br>
<br>

#### Exclude genes with mean TPM < 1
```{r filter_TPM, echo=TRUE, fig.margin=TRUE}

gene_ID_failed=gene_mean$geneID[gene_mean$mean_TPM<1]

brain_cortex_tmp_pheno_cov_TPM_filtered=brain_cortex_tmp_pheno_cov[,!names(brain_cortex_tmp_pheno_cov) %in%gene_ID_failed]
dim(brain_cortex_tmp_pheno_cov_TPM_filtered)
```


<br>
<br>


#### The function for report the association results 

```{r linear_regression_fun, echo=TRUE, fig.margin=TRUE}
library(RNOmni)
library(robustbase)

Phenotype_header=c("AGE","SEX")
Covariate_header=c(names(brain_cortex_tmp_pheno_cov_TPM_filtered)[grepl('Inferred',names(brain_cortex_tmp_pheno_cov_TPM_filtered))],"PC1","PC2","PC3","PC4","PC5")

lm_sex <- function(Expression_header) {
  # Function body
  # Perform operations using arguments and other statements
  block=brain_cortex_tmp_pheno_cov_TPM_filtered[,c(Expression_header,Phenotype_header,Covariate_header)]

  block <- block %>%
    mutate_all(as.numeric)
  
  block$SEX=as.factor(block$SEX)
  
  
  formula <- paste("RankNorm(", Expression_header, ") ~ .", sep = "")

# Running robust linear regression with a maximum of 2000 iterations
  lm_test <- lmrob(formula, data = block, k.max = 2000)
  
  if (class(lm_test) == "lmrob" && "SEX2" %in% rownames(summary(lm_test)$coefficients)) {
      # Return the coefficients summary for the 'SEX' variable
      return(summary(lm_test)$coefficients['SEX2', ])
    } else {
      return(c(NA,NA,NA,NA))
    }
}


```

<br>
<br>

#### use the function to test first 500 genes by multithreading 

```{r test_2000, echo=TRUE, fig.margin=TRUE}

library(doParallel)
library(robustbase)

# Set up parallel processing
num_cores <- detectCores()  # Detect available cores
cl <- makeCluster(num_cores)  # Create a cluster using available cores
registerDoParallel(cl)  # Register the cluster for parallel processing

# Extract gene list
gene_list <- names(brain_cortex_tmp_pheno_cov_TPM_filtered)[grepl("ENSG", names(brain_cortex_tmp_pheno_cov_TPM_filtered))]

# Subset gene list
gene_list <- gene_list

# Define function for parallel processing

# Perform parallel processing
results <- foreach(gene = gene_list, .combine = rbind, .packages = c("dplyr","RNOmni","robustbase")) %dopar% {
  lm_sex(gene)
}

# Organize results into a data frame
results <- data.frame(gene = gene_list,
                      results)

# Order results by p-value
results <- results[order(results$P),]

# Stop parallel processing
stopCluster(cl)

# View the top of the results
head(results)
```


#### visualize gene expression of ENSG00000157870.14 by SEX
```{r box_age_ENSG00000157870.14, echo=TRUE, fig.margin=TRUE}

library(ggplot2)

# Create a box plot using ggplot2
ggplot(brain_cortex_tmp_pheno_cov_TPM_filtered, aes(x = SEX, y = RankNorm(as.numeric(ENSG00000157870.14)), fill = as.factor(SEX))) +
  geom_boxplot() +
  labs(title = "Box Plot by SEX", x = "SEX", y = "ENSG00000157870.14") +
  theme_minimal()

```


#### write the results to the file
```{r save_results, echo=TRUE, fig.margin=TRUE}

results$FDR=p.adjust(results$Pr...t..,method = 'fdr')

write.table(results, "brain_cortex_lmrob_SEX_multhread_with_ranknorm_FDR.txt", quote=F, row.names = F, sep='\t')

```